---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(pedtools)
library(segregatr)
knitr::opts_chunk$set(
  echo = FALSE,
  dpi = 300,
  comment = NA
)
```

# shinyseg

> ***Try shinyseg here***: [https://chrcarrizosa.shinyapps.io/shinyseg](https://chrcarrizosa.shinyapps.io/shinyseg).

## What is shinyseg?
shinyseg is an interactive web application for clinical cosegregation analysis, a method used to ascertain disease-associated genetic variants from family data. More specifically, shinyseg calculates the full-likelihood Bayes factor (FLB) for a variant's pathogenicity ([Thompson et al., 2003](https://doi.org/10.1086/378100)), which can be utilized within the ACMG-AMP framework for clinical variant interpretation ([Richards et al., 2015](https://doi.org/10.1038%2Fgim.2015.30)).

Analyses with shinyseg are easy and highly flexible. All you need to do is upload a pedigree file, enter the genetic and clinical data, and specify a suitable penetrance model. The following sections briefly explain how to do so for an X-linked recessive case (`Example 2`), but you can find more detailed information and tips within the app.

The application is powered by the [ped suite](https://magnusdv.github.io/pedsuite/), [Shiny](https://shiny.posit.co/), and [rhandsontable](https://CRAN.R-project.org/package=rhandsontable).

## Uploading a pedigree

shinyseg works with pedigree files in *ped* format, like those created by [QuickPed](https://magnusdv.shinyapps.io/quickped). Here is an example of such a file and the corresponding family:

:::: {style="display:flex; height:250px;"}
::: {style="width: 35%;"}
```{r ped-empty, fig.height = 2.25, fig.width = 2.5, out.width="95%", out.height="100%", fig.align = "center"}
x = avuncularPed(type = "maternal") |>
  addChildren(fa = 4, mo = 5, verbose = FALSE) |>
  relabel("asPlot")
plot(
  x,
  symbolsize = .82,
  cex = 0.7
)
```
:::
::: {style="width:65%; margin-left:1.5rem; margin-top:1.5rem;"}
```{r}
df = as.data.frame(x)
print(df, row.names = FALSE)
```
:::
::::

Note that while other columns may be present, the application will only utilize: `id` (individual ID), `fid` (father's ID, 0 if not included in the pedigree), `mid` (mother's ID, 0 if not included in the pedigree) and `sex` (1 = male; 2 = female).

## Genetic and clinical data

Once a pedigree is loaded, users are presented with a spreadsheet-like table where they can enter the analysis data. This includes four columns:

* __phenotype:__ free-text field to specify disease phenotypes relevant to the analysis, e.g. `affected`, `breast cancer`, etc. The value `nonaff` is reserved for unaffected individuals. If empty, it indicates that the phenotype is unknown.
* __carrier:__ `neg` for non-carriers, `het` for heterozygous (or hemizygous) carriers, `hom` for homozygous carriers. If empty, it indicates that the carriership state is unknown.
* __proband:__ checkbox indicating the proband or index case.
* __age:__ integer between 1-100 specifying the age of disease onset or censoring.

Using the same example from before:

:::: {style="display:flex; height:250px;"}
::: {style="width: 35%;"}
```{r ped-filled, fig.height = 2.25, fig.width = 2.5, out.width="95%", out.height="100%", fig.align = "center"}
plotSegregation(
  x,
  affected = c(3, 6, 7),
  unknown = 1:2,
  carriers = c(3, 5, 6, 7),
  noncarriers = 4,
  proband = 6,
  fill = "gray30",
  labs = setNames(seq(pedsize(x)), c(80, 80, 40, 60, 60, 40, 40)),
  symbolsize = .82,
  cex = .7
)
```
:::
::: {style="width:65%; margin-left:1.5rem; margin-top:1.5rem;"}
```{r}
df2 =
  data.frame(
    df,
    phenotype = c("", "", "affected", "nonaff", "nonaff", "affected", "affected"),
    carrier = c("", "", "het", "neg", "het", "het", "het"),
    proband = c("", "", "", "", "", "X", ""),
    age = c(80, 80, 40, 60, 60, 40, 40)
  )
print(df2, row.names = FALSE)
```
:::
::::

## Penetrance model

The inheritance pattern combines choices for chromosome (`A`: autosomal, `X`: X-linked) and dominance (`D`: dominant, `R`: recessive, `I`: incomplete). In this example, we select X-linked recessive (`XR`), which will adjust the next steps accordingly.

Next, it is time to assign the probabilities of observing the analysis phenotypes conditional on carriership status (hereafter referred to as penetrances). This may be done in two ways:

### 1) Relative risk

This is a parametric version of the survival penetrances by [Belman et al. (2020)](https://doi.org/10.1038/s41436-020-0920-4). It is based on:

* __Baseline lifetime risk, mean and SD:__ the lifetime risk, mean, and standard deviation of disease onset in phenocopies, i.e., non-carriers and heterozygous carriers in recessive inheritance. 
* __Hazard ratios:__ the relative risks in homo-, hemi-, and heterozygous carriers in dominant inheritance, compared to phenocopies. They may also be specified through a __variant-associated lifetime risk__.

For example, the following defines that non-carrier men and heterozygous women have a lifetime risk of 0.01 (1%) with onset at 70±15 years of age, while hemizygous men and homozygous women have a lifetime risk of 0.75 (75%).

```{r}
df3 =
  data.frame(
    sex = c("", "sex", "both"),
    phenotype = c("", " phenotype", "affected"),
    f0R = c("  neg/♀het", "risk", "0.01"),
    f0mu = c("", " mu", "70"),
    f0sigma = c("", " sigma", "15"),
    f2R = c("  ♂het/hom", "risk", "0.75"),
    HR = c("", "hazard ratio(s)", "137.93")
  )
colnames(df3) = NULL
print(df3, row.names = FALSE, col.names = FALSE, right = TRUE)
```

Notice that the hazard ratio(s) column also indicates some value. In this case, it was automatically updated by shinyseg based on the other parameters, but it can be directly modified to specify more complex models. However, this may be difficult to do. For this reason, the app also features an assistant where users can instead enter cumulative incidence data, and the model parameters will be optimized. For instance:

:::: {style="display: flex;"}
::: {style="width: 37%;"}
```{r}
df4 =
  data.frame(
    age = c("", "age", "20", "40", "60", "80"),
    f0CI = c(" neg/♀het", "CI", "0.01", "0.02", "0.03", "0.04"),
    f2CI = c(" ♂het/hom", "CI", "0.01", "0.20", "0.60", "0.70")
  )
colnames(df4) = NULL
print(df4, row.names = FALSE, col.names = FALSE, right = TRUE)
```
:::
::: {style="width: 63%; margin-left:1rem;"}
```{r}
cat(
  "Optimal parameters:\n",
  "- neg/♀het risk: 0.0446\n",
  "- neg/♀het mean: 50.0096\n",
  "- neg/♀het SD: 20.0044\n",
  "- ♂het/hom risk: 0.7059\n",
  "- hazard ratio(s): 1.73, 1.87, 47.8, 51.47, 4.76, 2.74"
)
```
:::
::::

### 2) Liability class

Alternatively, it is possible to manually define penetrances using a table. The following defines the simplest case where there are no phenocopies, and the chance of disease onset in hemizygous men and homozygous women is 100%.

```{r}
df5 =
  data.frame(
    f0 = c("neg/♀het", "risk", "0"),
    f2 = c("♂het/hom", "risk", "1"),
    sex = c("", " sex", ""),
    phenotype = c("", " phenotype", ""),
    ages = c("", " ages", "")
  )
colnames(df5) = NULL
print(df5, row.names = FALSE, col.names = FALSE, right = TRUE)
```

A more detailed specification dependent on sex, phenotype, and age can be created by adding more rows.

## Calculating the FLB

Once all previous steps have been completed, obtaining the FLB is as easy as clicking on the corresponding button. For instance, assuming a variant population frequency of 0.001 and using the simple liability class model shown before, we obtain an `FLB = 8.00`. This value would indicate supporting evidence for pathogenicity according to [Jarvik and Browning's (2016)](https://doi.org/10.1016%2Fj.ajhg.2016.04.003) thresholds.

The application then opens up more possibilities, including downloading results in a custom HTML report or performing sensitivity analyses to examine the robustness of cosegregation evidence. Explore them now with [shinyseg](https://chrcarrizosa.shinyapps.io/shinyseg)!
