---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
library(pedtools)
library(segregatr)
knitr::opts_chunk$set(
  echo = FALSE,
  dpi = 300,
  comment = NA,
  fig.align = "center"
)
```

# shinyseg

> ***Try it here***: [https://chrcarrizosa.shinyapps.io/shinyseg](https://chrcarrizosa.shinyapps.io/shinyseg).

## What is shinyseg?

shinyseg is an interactive web application for clinical cosegregation analysis, a method used to ascertain disease-associated genetic variants from family data. More specifically, it calculates the full-likelihood Bayes factor (FLB) for a variant's pathogenicity ([Thompson et al., 2003](https://doi.org/10.1086/378100)), which can be readily used within the ACMG-AMP framework for clinical variant interpretation ([Richards et al., 2015](https://doi.org/10.1038%2Fgim.2015.30); [Jarvik and Browning, 2016](https://doi.org/10.1016%2Fj.ajhg.2016.04.003)).

Analyses with shinyseg are easy and highly flexible. All you need to do is upload a pedigree file, enter the genetic/clinical data, and specify a suitable penetrance model. The following sections briefly explain how to do so with an example case (`Example 2`), but you can find detailed guidance within the app.

shinyseg is powered by the [ped suite](https://magnusdv.github.io/pedsuite/), [Shiny](https://shiny.posit.co/), and [rhandsontable](https://CRAN.R-project.org/package=rhandsontable).

## Uploading a pedigree

The application works with pedigree files in ped format, like those created by [QuickPed](https://magnusdv.shinyapps.io/quickped). Here is an example of such a file and the corresponding family:

```{r}
x = avuncularPed(type = "maternal") |>
  addChildren(fa = 4, mo = 5, verbose = FALSE) |>
  relabel("asPlot")
df = as.data.frame(x)
print(df, row.names = FALSE)
```

```{r ped-empty, fig.height = 2.25, fig.width = 2.25, out.width="300px"}
plot(
  x,
  symbolsize = .82,
  cex = 0.7
)
```

Note that other columns may be present, but the app will only read: `id` (individual ID), `fid` (father's ID, 0 if not included in the pedigree), `mid` (mother's ID, 0 if not included in the pedigree), and `sex` (1 = male; 2 = female).

## Genetic and clinical data

Once a pedigree is loaded, users are presented with a table to enter the analysis data:

* __phenotype:__ free-text field to specify disease phenotypes relevant to the analysis, e.g. `affected`, `breast cancer`, etc. `nonaff` for unaffected individuals, empty if unknown.
* __carrier:__ `neg` for non-carriers, `het` for heterozygous (or hemizygous) carriers, `hom` for homozygous carriers. Empty if unknown.
* __proband:__ checkbox indicating the proband or index case.
* __age:__ integer between 1-100 specifying the age of disease onset or censoring.

For instance:

```{r}
df2 =
  data.frame(
    df,
    phenotype = c("", "", "affected", "nonaff", "nonaff", "affected", "affected"),
    carrier = c("", "", "het", "neg", "het", "het", "het"),
    proband = c("", "", "", "", "", "X", ""),
    age = c(80, 80, 40, 60, 60, 40, 40)
  )
print(df2, row.names = FALSE)
```

```{r ped-filled, fig.height = 2.25, fig.width = 2.25, out.width="300px"}
plotSegregation(
  x,
  affected = c(3, 6, 7),
  unknown = 1:2,
  carriers = c(3, 5, 6, 7),
  noncarriers = 4,
  proband = 6,
  fill = "gray30",
  labs = setNames(seq(pedsize(x)), c(80, 80, 40, 60, 60, 40, 40)),
  symbolsize = .82,
  cex = .7
)
```

## Penetrance model

The inheritance pattern combines choices for chromosome (`A`: autosomal, `X`: X-linked) and dominance (`D`: dominant, `R`: recessive, `I`: incomplete). In this example we select X-linked recessive (`XR`), which will adjust other inputs accordingly.

Next, it is time to assign the probabilities of observing the analysis phenotypes conditional on carriership status. These will be hereafter referred to as penetrances, and their specification may be done in two ways:

### 1) Relative risk

A parametric version of the survival penetrances by [Belman et al. (2020)](https://doi.org/10.1038/s41436-020-0920-4). It is based on:

* __Baseline lifetime risk, mean and SD:__ the lifetime risk, mean, and standard deviation of disease onset in phenocopies, i.e. non-carriers and heterozygous carriers in recessive inheritance. 
* __Hazard ratios:__ the relative risks in homo-, hemi-, and heterozygous carriers in dominant inheritance, compared to phenocopies. They can be either constant or age-dependent and may also be specified through a __variant-associated lifetime risk__.

For example, the following describes a constant relative risk model in which phenocopies have a lifetime risk of 0.01 (1%) with onset at 70±15 years of age, while hemizygous men and homozygous women have a lifetime risk of 0.75 (75%).

```{r}
df3 =
  data.frame(
    sex = c("", "sex", "both"),
    phenotype = c("", " phenotype", "affected"),
    f0R = c("  neg/♀het", "risk", "0.01"),
    f0mu = c("", " mu", "70"),
    f0sigma = c("", " sigma", "15"),
    f2R = c("  ♂het/hom", "risk", "0.75"),
    HR = c("hazard", "    ratio(s)", "137.93")
  )
colnames(df3) = NULL
print(df3, row.names = FALSE, col.names = FALSE, right = TRUE)
```

To facilitate the specification, shinyseg can also optimize these parameters based on user-provided cumulative incidence data.

### 2) Liability class

Alternatively, users can manually define the penetrances using a table. The following represents the simplest case where there are no phenocopies, and the chance of disease onset in hemizygous men and homozygous women is 100%.

```{r}
df4 =
  data.frame(
    f0 = c("neg/♀het", "risk", "0.00"),
    f2 = c("♂het/hom", "risk", "1.00"),
    sex = c("", " sex", ""),
    phenotype = c("", " phenotype", ""),
    ages = c("", " ages", "")
  )
colnames(df4) = NULL
print(df4, row.names = FALSE, col.names = FALSE, right = TRUE)
```

More detailed specifications, dependent on sex, phenotype, and age, can be created by adding more rows and filling in these columns accordingly.

## Cosegregation evidence

Calculating the FLB is then as easy as clicking a button. For instance, with a population variant frequency of `0.001` and the simple liability class model from before, we get an `FLB = 8.00`. shinyseg reports this as supporting evidence for pathogenicity based on [Jarvik and Browning's (2016)](https://doi.org/10.1016%2Fj.ajhg.2016.04.003) thresholds.

Afterward, the app opens up more possibilities, including downloading a report with your results or performing sensitivity analyses to assess their robustness. Explore all these features now with [shinyseg](https://chrcarrizosa.shinyapps.io/shinyseg)!
