---
output: html_fragment
---

<!-- pedigreeBox.html is generated from pedigreeBox.Rmd. Please edit that file -->

```{r, include = FALSE}
library(pedtools)
library(segregatr)
knitr::opts_chunk$set(
  echo = FALSE,
  dpi = 300,
  comment = NA,
  fig.align = "center"
)
```

**Uploading a pedigree**

shinyseg works with pedigree files in ped format. These have a tabular structure with columns:

-   `id`: individual ID.
-   `fid`: father's ID, 0 if not included in the pedigree.
-   `mid`: mother's ID, 0 if not included in the pedigree.
-   `sex`: 1 = male; 2 = female.
    (Note: Additional columns may be present, but the app will only read these.)

Here is an example of such a file and the corresponding family:

:::: {style="display:flex;"}

::: {}
```{r}
x = avuncularPed(type = "maternal") |>
  addChildren(fa = 4, mo = 5, verbose = FALSE) |>
  relabel("asPlot")
df = as.data.frame(x)
print(df, row.names = FALSE)
```
:::

::: {style="margin-left: 2rem;"}
```{r ped-empty, fig.height = 2.25, fig.width = 2.25, out.width="225px", out.height="225px"}
plot(
  x,
  symbolsize = .82,
  cex = 0.7
)
```
:::

::::

If you need help creating the ped file, you can [<i class='fa-solid fa-globe fa-fw'></i>Use QuickPed]{style="color:#4e4e4e;"}, which allows you to do so in an interactive way. Once ready, upload it by clicking on [<i class='fa-solid fa-file-import fa-fw'></i>Add pedfile]{style="color:#4e4e4e;"}. Alternatively, the dropdown menu on the left offers some basic structures that may suit your analysis.

You can repeat this process to add more families; they will appear concatenated in the table below and indexed by the `ped` column. Alternatively, you may upload multiple families simultaneously by providing this column yourself. To remove a family, click on [<i class='fa-solid fa-person-digging fa-fw'></i>Modify]{style="color:#4e4e4e;"} > [<i class='fa-solid fa-users-slash fa-fw'></i>]{style="color:#4e4e4e;"}.

> <i class='fa fa-triangle-exclamation fa-fw' style='color:#db1f48'></i> You will see a warning until you add at least one family to the analysis.

<br>

**Filling in the data**

Upon loading a pedigree, users are presented with a table to enter the clinical and genetic data. This consists of four additional columns:

-   **phenotype:** a free-text field to specify disease phenotypes relevant to the analysis, e.g., `affected`, `breast cancer`, etc. Use `nonaff` for unaffected individuals; leave empty if unknown.
-   **carrier:** use `neg` for non-carriers, `het` for heterozygous (or hemizygous) carriers, `hom` for homozygous carriers. Leave empty if unknown.
-   **proband:** a checkbox indicating the proband or index case.
-   **age:** an integer between 1-100 specifying the age of disease onset or censoring/last follow-up.

For instance, the following states that individuals 3, 6, and 7 had the onset of a certain disease at 40 years of age. Conversely, 4 and 5 were unaffected by 60 years of age. Individuals 3, 5, 6 (proband), and 7 carry one copy of the variant under study; 4 does not. For the founders 1 and 2, only ages are known:

::: {style="display:flex;"}
<div>

```{r}
df2 =
  data.frame(
    df,
    phenotype = c("", "", "affected", "nonaff", "nonaff", "affected", "affected"),
    carrier = c("", "", "het", "neg", "het", "het", "het"),
    proband = c("", "", "", "", "", "X", ""),
    age = c(80, 80, 40, 60, 60, 40, 40)
  )
print(df2, row.names = FALSE)
```

</div>

::: {style="margin-left: 2rem;"}
```{r, fig.height = 2.25, fig.width = 2.25, out.width="225px", out.height="225px"}
plotSegregation(
  x,
  affected = c(3, 6, 7),
  unknown = 1:2,
  carriers = c(3, 5, 6, 7),
  noncarriers = 4,
  proband = 6,
  fill = "gray30",
  labs = setNames(seq(pedsize(x)), c(80, 80, 40, 60, 60, 40, 40)),
  symbolsize = .82,
  cex = .7
)
```
:::
:::

Note that you can also upload these data together with your pedigree via [<i class='fa-solid fa-file-import fa-fw'></i>Add pedfile]{style="color:#4e4e4e;"}. To ensure smooth processing, we recommend replacing the empty cells with dots when doing so.

> <i class='fa fa-triangle-exclamation fa-fw' style='color:#db1f48'></i> For each family, ensure you have: 1 or more disease phenotypes, 1 or more variant carriers, 1 proband (who should have both disease and variant), and no missing ages.
