---
output: html_fragment
---

<!-- penetranceBox.html is generated from penetranceBox.Rmd. Please edit that file -->

```{r, include = FALSE}
library(pedtools)
library(segregatr)
knitr::opts_chunk$set(
  echo = FALSE,
  dpi = 300,
  comment = NA,
  fig.align = "center"
)
```

**The inheritance pattern**

Choose the inheritance pattern using the buttons on the right of the header:

:::: {style="display:flex;"}

::: {}
-   `AD`: autosomal dominant.
-   `AR`: autosomal recessive.
-   `AI`: autosomal incomplete dominance.
:::

::: {}
-   `XD`: X-linked dominant.
-   `XR`: X-linked recessive.
-   `XI`: X-linked incomplete dominance.
:::

::::

Note that incomplete dominance, where you can specify different parameters for the carriers 1 and 2 copies of the variant, will be unavailable unless you also switch to the `Liability class` mode.

<br>

**Specifying the penetrances**

shinyseg offers two different ways for specifying the penetrances, or probabilities of disease onset conditional on carriership status.

- `Relative risk`, where the penetrances are computed by the app based on certain parameters.
- `Liability class`, where the user manually specifies the penetrances.

The modes can be switched using the input on the left of the header.

<br>

**Mode: Relative risk**

A parametric version of the survival penetrances introduced by [Belman et al. (2020)](https://doi.org/10.1038/s41436-020-0920-4). It is based on:

-   __Baseline lifetime risk, mean and SD:__ the lifetime risk, mean, and standard deviation of disease onset in non-carriers and heterozygous carriers in recessive inheritance.
-   __Hazard ratios:__ the relative risks in homo-, hemi-, and heterozygous carriers in dominant inheritance, compared to the baseline. They can be either constant or age-dependent and may also be specified through a __variant-associated lifetime risk__.

These parameters are entered using a table generated from the phenotypes specified in the pedigree table. For example, the following describes a constant relative risk model in which non-carriers have a lifetime risk of `affected` of 1% (phenocopy rate) with onset at 70±15 years of age, while heterozygous and homozygous carriers have a lifetime risk of 75%.

```{r}
df1 =
  data.frame(
    sex = c("sex", "both"),
    phenotype = c(" phenotype", "affected"),
    f0R = c("risk", "0.01"),
    f0mu = c(" mean", "70"),
    f0sigma = c(" SD", "15"),
    f2R = c("risk", "0.75"),
    HR = c("    hazard ratio(s)", "137.93")
  )
colnames(df1) = c("", "", "  neg", "", "", "  het/hom", "")
print(df1, row.names = FALSE, right = TRUE)
```

*By default, shinyseg generates two entries for each phenotype to allow for sex-specific penetrances. Changing either sex to `both` applies its parameters to both sexes. To return to sex-dependent specification, select one of the sexes again.

In the previous example, the hazard ratios were represented by a single value, indicating a constant (age-independent) relative risk. Additionally, this was automatically updated based on the other parameters. However, this column can also be directly manipulated and used to specify models where relative risks are age-dependent. To do this, enter up to 10 comma-separated values representing, loosely speaking, the hazard ratios at evenly spaced ages from 1 to 100 years. For instance, `1,10,10,20,10` would correspond to the ratios at ages 1, 25, 50, 75 and 100 (approximately). The app will interpolate/smooth these across the lifespan using a B-spline basis on the log scale — you may adjust the degree of smoothing with <span style='color:#4e4e4e;'><i class='fa-solid fa-wave-square fa-fw'></i>Splines</span>.

Note that it is also possible to leverage the variant-associated lifetime risk, as earlier, but combined with a vectorial specification of the hazard ratios. For instance, changing the hazard ratio(s) to `1,2,2,1` defines a model where the relative risk is higher at middle ages. Subsequently, resetting the lifetime risk to 75% will scale these values accordingly, resulting in:

```{r}
df2 =
  data.frame(
    sex = c("sex", "both"),
    phenotype = c(" phenotype", "affected"),
    f0R = c("risk", "0.01"),
    f0mu = c(" mean", "70"),
    f0sigma = c(" SD", "15"),
    f2R = c("risk", "0.75"),
    HR = c("    hazard ratio(s)", "    1, 305.02, 305.02, 1")
  )
colnames(df2) = c("", "", "  neg", "", "", "  het/hom", "")
print(df2, row.names = FALSE, right = TRUE)
```

As choosing these parameters can be challenging, shinyseg offers an <span style='color:#4e4e4e;'><i class='fa-solid fa-circle-info fa-fw'></i>Assistant</span> to optimize them based on user-provided cumulative incidence data. Additionally, the <span style='color:#4e4e4e;'><i class='fa-solid fa-chart-line fa-fw'></i>Plot</span> serves as a visual guide, displaying the cumulative incidences, hazard rates, or hazard ratios determined by the current selection.

> <i class='fa fa-triangle-exclamation fa-fw' style='color:#db1f48'></i> None of the model parameters can be missing. If you misspecify the hazard ratio(s), their values will revert to the previous ones to prevent issues.

<br>
<br>

**Mode: Liability class**

Alternatively, users can manually define penetrances using a table. This table contains 2 (dominant/recessive) to 3 (incomplete inheritance) columns with the penetrances themselves, followed by an additional 3 columns specifying the classes they apply to:

-   __sex:__ `male`, `female`, or `both` for classes referring to males, females, or both sexes, respectively; leaving this cell empty is equivalent to specifying `both`.
-   __phenotype:__ a free-text field whose values should match the phenotypes in the pedigree table. Note that here, `nonaff` is treated as another phenotype whose classes must be specified (unless there are no unaffected cases); leaving this cell empty indicates that the class applies to all phenotypes.
-   __ages:__ an integer from 1 to 100 or ranges between these values (e.g., 10-20, 30-100...) indicating the ages to which the class applies; leaving this cell empty is equivalent to specifying ages 1-100.

For instance, the following illustrates an autosomal dominant model with no phenocopies, where both heterozygous and homozygous carriers have a 100% chance of disease onset:

```{r}
df3 =
  data.frame(
    f0 = c("risk", "0.00"),
    f2 = c("risk", "1.00"),
    sex = c("  sex", ""),
    phenotype = c(" phenotype", ""),
    ages = c(" ages", "")
  )
colnames(df3) = c("neg", "  het/hom", "", "", "")
print(df3, row.names = FALSE, right = TRUE)
```

And the following defines a sex- and age-dependent specification with incomplete dominance:

```{r}
df4 =
  data.frame(
    f0 = c("risk", "0.00", "0.00", "0.05", "0.20"),
    f1 = c("risk", "0.20", "0.20", "0.40", "0.40"),
    f2 = c("risk", "0.40", "0.40", "0.70", "0.90"),
    sex = c("     sex", "male", "female", "male", "female"),
    phenotype = c("  phenotype", "", "", "", ""),
    ages = c("   ages", "1-50", "1-50", "51-100", "51-100")
  )
colnames(df4) =  c("neg", "  het", "  hom", "", "", "")
print(df4, row.names = FALSE, right = TRUE)
```


The table may be created interactively or by uploading a file via <span style='color:#4e4e4e;'><i class='fa-solid fa-file-import fa-fw'></i>Load file</span>. If opting for the latter, shinyseg expects a tabular file without column headers and a structure resembling the one shown in-app.

> <i class='fa fa-triangle-exclamation fa-fw' style='color:#db1f48'></i> There should be no sex/phenotype/age overlaps. All family members in the pedigree table should map to a single liability class without missing penetrances.
